---
title: "Exploring Storm Data"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE)
```

Introduction

Storms and other severe weather events can cause both public health and economic problems for communities and municipalities. Many severe events can result in fatalities, injuries, and property damage, and preventing such outcomes to the extent possible is a key concern.

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database. This database tracks characteristics of major storms and weather events in the United States, including when and where they occur, as well as estimates of any fatalities, injuries, and property damage.

Data

The data for this assignment come in the form of a comma-separated-value file compressed via the bzip2 algorithm to reduce its size. You can download the file from the course web site:

Storm Data [47Mb]
There is also some documentation of the database available. Here you will find how some of the variables are constructed/defined.

National Weather Service Storm Data Documentation

National Climatic Data Center Storm Events FAQ

The events in the database start in the year 1950 and end in November 2011. In the earlier years of the database there are generally fewer events recorded, most likely due to a lack of good records. More recent years should be considered more complete.

Assignment

The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.

Questions

Your data analysis must address the following questions:

Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

Across the United States, which types of events have the greatest economic consequences?

Consider writing your report as if it were to be read by a government or municipal manager who might be responsible for preparing for severe weather events and will need to prioritize resources for different types of events. However, there is no need to make any specific recommendations in your report.

Requirements

For this assignment you will need some specific tools

RStudio: You will need RStudio to publish your completed analysis document to RPubs. You can also use RStudio to edit/write your analysis.

knitr: You will need the knitr package in order to compile your R Markdown document and convert it to HTML

Document Layout

Language: Your document should be written in English.

Title: Your document should have a title that briefly summarizes your data analysis

Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the cache = TRUE option for certain code chunks.

There should be a section titled Results in which your results are presented.

You may have other sections in your analysis, but Data Processing and Results are required.

The analysis document must have at least one figure containing a plot.

Your analyis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.

You must show all your code for the work in your analysis document. This may make the document a bit verbose, but that is okay. In general, you should ensure that echo = TRUE for every code chunk (this is the default setting in knitr).

## Synopsis

## Data Processing

Load the necessary packages:

```{r}
library(dplyr)
library(ggplot2)
library(stringr)
```

Read the storm data, change the field names to lower case, subset the data to pull fields that matters, and convert event type values to lowercase.

```{r}
data <- read.csv("repdata-data-StormData.csv", stringsAsFactors = FALSE)
names(data) <- tolower(names(data))
subdata <- select(data, bgn_date, state, evtype, fatalities, injuries, propdmg, propdmgexp, cropdmg, cropdmgexp, remarks)
subdata$evtype <- tolower(subdata$evtype)
```

Check which field to clean up:

```{r}
test <- arrange(summarise(group_by(subdata, evtype), num.event = n()), evtype)
hail <- arrange(test[grep("hail", test$evtype),],desc(num.event))
thunderstorm <- arrange(test[grep("tstm", test$evtype),],desc(num.event))
```

Scrub data to clean up event type field.

```{r}
subdata$evtype <- str_trim(subdata$evtype, side = "both")
subdata$evtype <- gsub("^tstmw$", "thunderstorm wind", subdata$evtype)
subdata$evtype <- gsub("^tstm.*", "thunderstorm", subdata$evtype)
subdata$evtype <- gsub("^thu.*", "thunderstorm wind", subdata$evtype)
subdata$evtype <- gsub("^flash.*", "flash flood", subdata$evtype)
subdata$evtype <- gsub("^hail.*", "hail", subdata$evtype)
subdata$evtype <- gsub("^high w.*", "high wind", subdata$evtype)
subdata$evtype <- gsub("^heavy sn.*", "heavy snow", subdata$evtype)
subdata$evtype <- gsub("^heavy r.*", "heavy rain", subdata$evtype)
subdata$evtype <- gsub("^heavy su.*", "heavy surf", subdata$evtype)
subdata$evtype <- gsub(".*floodin.*", "flood", subdata$evtype)
subdata$evtype <- gsub("^flood.*", "flood", subdata$evtype)
subdata$evtype <- gsub("^summary.*", "summary (ignore)", subdata$evtype)
subdata$evtype <- gsub("^rip c.*", "rip current", subdata$evtype)
subdata$evtype <- gsub("^lightning.*", "lightning", subdata$evtype)
subdata$evtype <- gsub("^hurricane.*", "hurricane", subdata$evtype)
subdata$evtype <- gsub("^typhoon.*", "typhoon", subdata$evtype)
subdata$evtype <- gsub("^torna.*", "tornado", subdata$evtype)
subdata$evtype <- gsub("^watersp.*", "waterspout", subdata$evtype)
subdata$evtype <- gsub("^frost.*", "frost/freeze", subdata$evtype)
subdata$evtype <- gsub("^freeze.*", "frost/freeze", subdata$evtype)
subdata$evtype <- gsub("^tropical s.*", "tropical storm", subdata$evtype)
subdata$evtype <- gsub("^coastal.*", "coastal flood", subdata$evtype)
subdata$evtype <- gsub("^blizzard.*", "blizzard", subdata$evtype)
subdata$evtype <- gsub("^volcanic.*", "volcanic ash", subdata$evtype)
subdata$evtype <- gsub("^wild.*", "wildfire", subdata$evtype)
subdata$evtype <- gsub("^beach.*", "coastal flood", subdata$evtype)
subdata$evtype <- gsub("^avalan.*", "avalanche", subdata$evtype)
subdata$evtype <- gsub("^funnel.*", "funnel cloud", subdata$evtype)
subdata$evtype <- gsub("^freezing fog$", "fog", subdata$evtype)
subdata$evtype <- gsub("^freezing.*$", "frost/freeze", subdata$evtype)
subdata$evtype <- gsub("^fog$", "freezing fog", subdata$evtype)
subdata$evtype <- gsub("^unseason.*warm.*", "heat", subdata$evtype)
subdata$evtype <- gsub("^unseason.*co.*", "cold/wind chill", subdata$evtype)
subdata$evtype <- gsub("^unseason.*rain.*|^unseason.*wet.*", "heavy rain", subdata$evtype)
subdata$evtype <- gsub("^unseason.*hot.*|^unseason.*dry.*", "heat", subdata$evtype)
subdata$evtype <- gsub("^unseason.*low temp.*", "cold/wind chill", subdata$evtype)
subdata$evtype <- gsub("^ext.*co.*|^exc.*co.*|^ext.*wind.*|^record.*co.*|^record.*co.*|^record low$|^unusual.*cold", "extreme cold/wind chill", subdata$evtype)
subdata$evtype <- gsub("^ext.*he.*|^exc.*he.*|^record.*hi.*|^record.*he.*|^record.*warm.*|^unusual.*warmth.*|^very warm$", "excessive heat", subdata$evtype)
subdata$evtype <- gsub("^exc.*prec.*|^exc.*rai.*|^exc.*wet.*|^ext.*wet.*|^record.*prec.*|^record.*rai.*|^torr.*rai.*", "heavy rain", subdata$evtype)
subdata$evtype <- gsub("^exc.*dry.*|^record.*dry.*|^very.*dry.*", "drought", subdata$evtype)
subdata$evtype <- gsub("^exc.*snow.*|^record.*snow.*|^unusual.*snow.*", "heavy snow", subdata$evtype)
subdata$evtype <- gsub("^sleet.*", "sleet", subdata$evtype)
subdata$evtype <- gsub(".*microb.*|.*mircob.*|.*micob.*", "thunderstorm wind", subdata$evtype)
subdata$evtype <- gsub("^dry.*|^driest.*|^drou.*", "drought", subdata$evtype)
subdata$evtype <- gsub("^sno.*", "winter weather", subdata$evtype)
subdata$evtype <- gsub("^winter storm.*", "winter storm", subdata$evtype)
subdata$evtype <- gsub("^winter.*mix$|^winter weather.*|^wintry.*", "winter weather", subdata$evtype)
subdata$evtype <- gsub("^urban.*|^small stream.*|^sml stream.*", "flood", subdata$evtype)
subdata$evtype <- gsub("^blowing sn.*", "heavy snow", subdata$evtype)
subdata$evtype <- gsub("^light sn.*", "winter weather", subdata$evtype)
subdata$evtype <- gsub("^heavy precip.*", "heavy rain", subdata$evtype)
subdata$evtype <- gsub("^heat.*", "heat", subdata$evtype)
subdata$evtype <- gsub("^dust dev.*", "dust devil", subdata$evtype)
subdata$evtype <- gsub("^dust st.*|^dusts.*", "dust storm", subdata$evtype)
subdata$evtype <- gsub("^gusty w.*", "strong wind", subdata$evtype)
subdata$evtype <- gsub("^gusty thunder.*", "thunderstorm wind", subdata$evtype)
subdata$evtype <- gsub("^gustna.*", "tornado", subdata$evtype)
subdata$evtype <- gsub("^wind chill.*", "cold/wind chill", subdata$evtype)
subdata$evtype <- gsub("^wind.*", "high wind", subdata$evtype)
subdata$evtype <- gsub(".*lake.*snow.*", "lake-effect snow", subdata$evtype)
subdata$evtype <- gsub("^gusty lake wind$", "seche", subdata$evtype)
subdata$evtype <- gsub("^lake.*flood.*", "lakeshore flood", subdata$evtype)
subdata$evtype <- gsub("^severe cold.*", "extreme cold/wind chill", subdata$evtype)
subdata$evtype <- gsub("^fog and cold.*", "freezing fog", subdata$evtype)
subdata$evtype <- gsub("^cold.*", "cold/wind chill", subdata$evtype)
subdata$evtype <- gsub("^rain.*|^prolonged rain$|^early rain$|^monthly rainfall$|^hvy rain$|^locally heavy rain$", "heavy rain", subdata$evtype)
subdata$evtype <- gsub("^light freezing rain$", "winter weather", subdata$evtype)
subdata$evtype <- gsub(".*ice.*storm.*", "ice storm", subdata$evtype)
subdata$evtype <- gsub("^ice storm$", "storm storm storm", subdata$evtype)
subdata$evtype <- gsub(".*ice.*|.*icy.*", "winter weather", subdata$evtype)
subdata$evtype <- gsub("storm storm storm", "ice storm", subdata$evtype)
subdata$evtype <- gsub(".*landslide.*|.*muds.*|.*mud.*slide.*", "debris flow", subdata$evtype)
subdata$evtype <- gsub("^strong win.*", "strong wind", subdata$evtype)
subdata$evtype <- gsub("^erosion/cstl flood.*", "coastal flood", subdata$evtype)
subdata$evtype <- gsub("^local flash flood.*", "flash flood", subdata$evtype)
subdata$evtype <- gsub("^local flood.*|^m.*flood.*|^r.*flood.*|^street flood.*|^tidal flood.*", "flood", subdata$evtype)
subdata$evtype <- gsub("^deep hail.*|^late season hail.*|^non severe hail.*|^small hail.*", "hail", subdata$evtype)
subdata$evtype <- gsub("^storm surge.*", "storm surge/tide", subdata$evtype)
subdata$evtype <- gsub("^severe thun.*", "tornado", subdata$evtype)
subdata$evtype <- gsub(".*frost$", "frost/freeze", subdata$evtype)
subdata$evtype <- gsub(".*freeze$", "frost/freeze", subdata$evtype)
subdata$evtype <- gsub(".*warm.*", "heat", subdata$evtype)
subdata$evtype <- gsub(".*wet snow.*", "sleet", subdata$evtype)
subdata$evtype <- gsub("^early snow.*|^acc.*snow.*|^dri.*snow.*|^fir.*snow.*|^lac.*snow.*|^lat.*snow.*|^mo.*snow.*|^near.*snow.*|^pro.*snow.*|^sea.*snow.*", "winter weather", subdata$evtype)
subdata$evtype <- gsub("^cool.*", "cold/wind chill", subdata$evtype)
subdata$evtype <- gsub("^hot.*", "heat", subdata$evtype)
subdata$evtype <- gsub("^hyp.*thermia.*", "cold/wind chill", subdata$evtype)
subdata$evtype <- gsub("^^high sur.*|high sea.*|^high.*sw.*|^high tides$", "high surf", subdata$evtype)
subdata$evtype <- gsub("^high.*win.*", "high wind", subdata$evtype)
subdata$evtype <- gsub("^high temp.*", "excessive heat", subdata$evtype)
subdata$evtype <- gsub("^wa.*spout.*", "waterspout", subdata$evtype)
subdata$evtype <- gsub("^wet.*", "heavy rain", subdata$evtype)
subdata$evtype <- gsub("^torn.*", "tornado", subdata$evtype)
subdata$evtype <- gsub("^smoke.*", "dense smoke", subdata$evtype)
subdata$evtype <- gsub("^lig.*", "lightning", subdata$evtype)
subdata$evtype <- gsub("^low.*", "cold/wind chill", subdata$evtype)
subdata$evtype <- gsub("^bitter win.*", "extreme cold/wind chill", subdata$evtype)
subdata$evtype <- gsub("^gradient win.*", "rip current", subdata$evtype)
subdata$evtype <- gsub("^glaze.*", "freezing fog", subdata$evtype)
subdata$evtype <- gsub("^heavy sho.*", "heavy rain", subdata$evtype)
subdata$evtype <- gsub("^brush.*|^grass fires$", "wildfire", subdata$evtype)
```

Add event year to the dataset and discard data prior to 1996. This is because the Storm Database has complete data for all events that occured from 1996 onwards. See the page [here] (https://www.ncdc.noaa.gov/stormevents/details.jsp) 

```{r}
subdata <- mutate(subdata, date = as.Date(data$bgn_date, "%m/%d/%Y %H:%M:%S"))
subdata <- mutate(subdata, event_year = year(date))
subdata <- filter(subdata, event_year >= 1996)
```

Add fields to get the true value of property and crop damage.

```{r}
subdata <- mutate(subdata, propdmgtotal = ifelse(propdmgexp == "K", propdmg * 1000, ifelse(propdmgexp == "M", propdmg * 1000000, ifelse(propdmgexp == "B",propdmg * 1000000000, propdmg *1))))
subdata <- mutate(subdata, cropdmgtotal = ifelse(cropdmgexp == "K", cropdmg * 1000, ifelse(cropdmgexp == "M", cropdmg * 1000000, ifelse(cropdmgexp == "B",cropdmg * 1000000000, cropdmg *1))))
subdata <- mutate(subdata, totaldmg = propdmgtotal + cropdmgtotal)
```

Create subsets of data for economic damages, fatalities, and injuries

```{r}
totaldmg <- arrange(top_n(summarise(group_by(subdata, evtype), total_damage = sum(totaldmg)), 10), desc(total_damage))
fatalities <- arrange(top_n(summarise(group_by(subdata, evtype), fatalities = sum(fatalities)), 10), desc(fatalities))
injuries <- arrange(top_n(summarise(group_by(subdata, evtype), injuries = sum(injuries)), 10), desc(injuries))
```

## Results

Natural storms have tremendous economic consequences. We aggregated total property and crop damages between 1996 to 2011 in the United States and came up with the top 10 storm types. Flood tops the list with over $150 billion in economic damages over the 15-year period, followed by hurricanes, and storm surge.

```{r}
ggplot(data = totaldmg, aes(x = reorder(evtype, total_damage), y = total_damage)) + geom_bar(stat = "identity", fill = "#F8766D") + coord_flip() + ggtitle("Top 10 Property and Crop Damage by Storm Type") + ylab("Total Damage (in million dollars)") + xlab("Storm Type")
```

As for health consequences, I looked at both fatalities and injuries. Data has suggested that excessive heat and tornadoes caused the most deaths with approximately 1,800 and 1,500 people respectively. 

```{r}
ggplot(data = fatalities, aes(x = reorder(evtype, fatalities), y = fatalities)) + geom_bar(stat = "identity", fill = "#00BF7D") + coord_flip() + ggtitle("Top 10 Fatalities by Storm Type") + ylab("Total Fatalities") + xlab("Storm Type")
```

In addition to deaths, tornadoes also topped the top 10 injuries list. It caused the most injuries by far compared to the rest on this list. Flood is second and excessive heat is third with a little over 6,000 in the 15-year period.